# Time course analysis

Want to compare dynamics of OTUs from the delta Met and detla Vanc treatments
over the course of the model. Why? These were the communities that were
treated with antibiotics and moved the most. The delta Met mice had between
5 and 10 tumors while the delta Vanc mice had 0 or 1 tumor.


```{r setup, results = "hold", message=FALSE}
library(randomForest, quietly=TRUE)
library(knitr, quietly=TRUE)
sessionInfo()

set.seed(6201976)

source("code/rf_baseline_analysis.R")

opts_chunk$set(dev = c("png", "cairo_pdf"))
opts_chunk$set(results = "hold")
opts_chunk$set(fig.show = "hold")
opts_chunk$set(warning = FALSE)
opts_chunk$set(fig.align = "center")
opts_chunk$set(echo = FALSE)
opts_chunk$set(cache = FALSE)
```


Would like to generate a plot over time indicating change in a few OTUs in
both conditions over the course of the model.

```{r get_data}
source("code/rf_baseline_analysis.R")

shared_file <- "data/process/ab_aomdss.trim.contigs.good.unique.good.filter.unique.precluster.pick.pick.pick.an.unique_list.0.03.subsample.shared"
rel_abund <- get_rel_abund(paste0(PROJHOME, "/", shared_file), "DSS")
rel_abund <- rel_abund[!grepl("dneg", rownames(rel_abund)),]

del_met <- "VancStrep"
del_met_rabund <- rel_abund[grep(del_met, rownames(rel_abund)),]

del_vanc <- "StrepMetro"
del_vanc_rabund <- rel_abund[grep(del_vanc, rownames(rel_abund)),]

del_rabund <- rbind(del_met_rabund, del_vanc_rabund)

del_days <- sapply(rownames(del_rabund), get_day)
del_days[del_days==70] <- 73
del_treat <- sapply(rownames(del_rabund), get_treatment)


otu03 <- aggregate(del_rabund$Otu00003, by=list(del_days, del_treat), summary)
del_met_otu03 <- otu03[otu03$Group.2 == del_met,]
del_vanc_otu03 <- otu03[otu03$Group.2 == del_vanc,]

otu06 <- aggregate(del_rabund$Otu00006, by=list(del_days, del_treat), summary)
del_met_otu06 <- otu06[otu06$Group.2 == del_met,]
del_vanc_otu06 <- otu06[otu06$Group.2 == del_vanc,]


jitter <- 0.5

par(mar=c(0.5,3,0.5,0.5), mfrow=c(2,1))
plot(NA, ylim=c(0, 0.75), xlim=c(0,75), xlab="", ylab="", axes=F)
polygon(c(5,5,10,10),c(80,-10,-10,80), col='light blue', border=NA)   # DSS 1 box
polygon(c(26,26,31,31),c(80,-10,-10,80), col='light blue', border=NA) # DSS 2 box
polygon(c(47,47,52,52),c(80,-10,-10,80), col='light blue', border=NA) # DSS 3 box

segments(x0=del_met_otu03$Group.1+jitter, y0=del_met_otu03$x[,"Min."],
            y1=del_met_otu03$x[,"Max."], lwd=2, col=clrs[del_met_otu03$Group.2])
points(del_met_otu03$Group.1+jitter, del_met_otu03$x[,"Median"], col=clrs[del_met_otu03$Group.2], type="l")
points(del_met_otu03$Group.1+jitter, del_met_otu03$x[,"Median"], pch=21, col=clrs[del_met_otu03$Group.2], bg="white", cex=1.5)

segments(x0=del_vanc_otu03$Group.1-jitter, y0=del_vanc_otu03$x[,"Min."],
            y1=del_vanc_otu03$x[,"Max."], lwd=2, col=clrs[del_vanc_otu03$Group.2])
points(del_vanc_otu03$Group.1-jitter, del_vanc_otu03$x[,"Median"], col=clrs[del_vanc_otu03$Group.2], type="l")
points(del_vanc_otu03$Group.1-jitter, del_vanc_otu03$x[,"Median"], pch=21, col=clrs[del_vanc_otu03$Group.2], bg="white", cex=1.5)

axis(2, at=seq(0,0.7,0.1), label=seq(0,70,10), las=2)
text(x=78, y=0.75, label="Enterobacteriaceae (OTU 3)", pos=2, font=2)
box()


plot(NA, ylim=c(0, 0.75), xlim=c(0,75), xlab="", ylab="", axes=F)
polygon(c(5,5,10,10),c(80,-10,-10,80), col='light blue', border=NA)   # DSS 1 box
polygon(c(26,26,31,31),c(80,-10,-10,80), col='light blue', border=NA) # DSS 2 box
polygon(c(47,47,52,52),c(80,-10,-10,80), col='light blue', border=NA) # DSS 3 box

segments(x0=del_met_otu06$Group.1+jitter, y0=del_met_otu06$x[,"Min."],
            y1=del_met_otu06$x[,"Max."], lwd=2, col=clrs[del_met_otu06$Group.2])
points(del_met_otu06$Group.1+jitter, del_met_otu06$x[,"Median"], col=clrs[del_met_otu06$Group.2], type="l")
points(del_met_otu06$Group.1+jitter, del_met_otu06$x[,"Median"], pch=21, col=clrs[del_met_otu06$Group.2], bg="white", cex=1.5)

segments(x0=del_vanc_otu06$Group.1-jitter, y0=del_vanc_otu06$x[,"Min."],
            y1=del_vanc_otu06$x[,"Max."], lwd=2, col=clrs[del_vanc_otu06$Group.2])
points(del_vanc_otu06$Group.1-jitter, del_vanc_otu06$x[,"Median"], col=clrs[del_vanc_otu06$Group.2], type="l")
points(del_vanc_otu06$Group.1-jitter, del_vanc_otu06$x[,"Median"], pch=21, col=clrs[del_vanc_otu06$Group.2], bg="white", cex=1.5)

axis(2, at=seq(0,0.7,0.1), label=seq(0,70,10), las=2)
text(x=78, y=0.75, label="Lactobacillus (OTU 6)", pos=2, font=2)
box()
```
