# Time course analysis


**Purpose:** We want to characterize the change in the microbiome over the
course of the model


### Setup

```{r setup, results = "hold", message=FALSE}
library(randomForest, quietly=TRUE)
library(knitr, quietly=TRUE)
sessionInfo()

set.seed(6201976)

source("code/rf_baseline_analysis.R")

```

```{r knitr_settings, echo=FALSE, eval=TRUE}
opts_chunk$set(dev = c("png", "cairo_pdf"))
opts_chunk$set(results = "hold")
opts_chunk$set(fig.show = "hold")
opts_chunk$set(warning = FALSE)
opts_chunk$set(fig.align = "center")
opts_chunk$set(echo = FALSE)
opts_chunk$set(cache = FALSE)
```

First, we want to read in the distance matrix and see which treatments are
changing the most...

```{r start_to_end}

get_dist_lt <- function(file_name){

    dist_vector <- scan(file_name, what="", sep="\n")

    n_samples <- as.numeric(dist_vector[1])
    dist_vector <- dist_vector[-1]

    sample_names <- character(n_samples)
    distance_matrix <- matrix(rep(0, n_samples*n_samples), nrow=n_samples)

    for(i in 1:n_samples){

        split_line <- unlist(strsplit(dist_vector[i], "\t"))
        sample_names[i] <- split_line[1]

        if(i > 1){
            distance_matrix[i,1:(i-1)] <- as.numeric(split_line[2:i])
        }
    }

    distance_matrix <- distance_matrix + t(distance_matrix)
    rownames(distance_matrix) <- sample_names
    colnames(distance_matrix) <- sample_names
    distance_matrix
}

get_start_end_dist <- function(u_treat_et){
    subset <- day[treat_et == u_treat_et]
    start <- names(subset[subset==0])
    end <- names(sort(subset, decreasing=TRUE))[1]
    distances[start, end]
}


dist_file <- "data/process/ab_aomdss.trim.contigs.good.unique.good.filter.unique.precluster.pick.pick.pick.an.unique_list.thetayc.0.03.lt.ave.dist"

distances <- get_dist_lt(dist_file)

treat_et <- gsub("_d.*", "", rownames(distances))
day <- sapply(rownames(distances), get_day)
eartag <- sapply(rownames(distances), get_eartag)

is_dss_only <- grepl("DSS", treat_et)
treat_et <- treat_et[!is_dss_only]
day <- day[!is_dss_only]
eartag <- eartag[!is_dss_only]
distances <- distances[!is_dss_only, !is_dss_only]

start_end_dist <- unlist(sapply(unique(treat_et), get_start_end_dist))
treatment <- sapply(names(start_end_dist), get_treatment)

aggregate(start_end_dist, by=list(treatment), summary)
```

Was surprised that the "AllAbs" seemed bimodal. Of the mice, 5 didn't move at
all and 3 moved a fair amount. Curious to see how well they replicated each other
at the beginning and the end of the model.


```{r compare_start}

get_start_dists <- function(single_treat){
    subset <- day[treatment==single_treat]
    start <- names(subset[subset==0])
    start_dists <- distances[start, start]
    summary(start_dists[lower.tri(start_dists)])
}

treatment <- sapply(rownames(distances), get_treatment)
uniq_treatments <- levels(factor(treatment))

sapply(uniq_treatments, get_start_dists)
```

Again, I was struck by the variation among the Day 0 samples for all
antibiotics. Let's see what it looks like at the end of the model...

```{r compare_end}

get_end_dists <- function(single_treat){
    subset_day <- day[treatment==single_treat]
    subset_et <- eartag[treatment==single_treat]

    terminal <- aggregate(subset_day, by=list(subset_et), max)

    end <- paste0(single_treat, "_", terminal$Group.1, "_d", terminal$x)
    end_dists <- distances[end, end]
    summary(end_dists[lower.tri(end_dists)])
}

sapply(uniq_treatments, get_end_dists)

```

More variation at the end point.

What happens when we plot the distance changed between day 0 and the end with
the number of tumors that were observed?


```{r distance_tumor_corr}
tumor_count.file <- paste0("data/process/tumor_counts.tsv")
tumor_counts <- read.table(file=tumor_count.file, header=T)

tumor_counts_prepost <- tumor_counts[names(start_end_dist)]
dist_tumor_cor <- cor.test(tumor_counts_prepost, start_end_dist, method="spearman")

treatments <- gsub("_.*", "", names(start_end_dist))
plot(tumor_counts_prepost ~start_end_dist, col=clrs[treatments],
    pch=pch[treatments], ylab="Number of tumors",
    xlab="Distance between start and end")
```

My eye wants to see some type of saturation curve where the more distance
between the start and end point, the more tumors occur. But... the $\rho$
statistic was `r round(dist_tumor_cor$estimate, 2)` (P=
`r round(dist_tumor_cor$p.value, 2)`).

Let's break down and build an ordination...

```{r}
get_start_sample <- function(treat_et){
    treat_et_days <- day[grep(treat_et, names(day))]
    names(treat_et_days[treat_et_days==0])
}

get_end_sample <- function(treat_et){
    treat_et_days <- day[grep(treat_et, names(day))]
    names(which.max(treat_et_days))
}

get_start_end_sample <- function(treat_et){
    c(get_start_sample(treat_et), get_end_sample(treat_et))
}

unique_treat_et <- unique(treat_et)

starts <- sapply(unique_treat_et, get_start_sample)
starts[sapply(starts, length)==0] <- NA
starts <- unlist(starts)

ends <- sapply(unique_treat_et, get_end_sample)
ends[sapply(ends, length)==0] <- NA
ends <- unlist(ends)

starts_ends <- c(starts, ends)
starts_ends <- starts_ends[!is.na(starts_ends)]

write(file="data/process/start_end_samples.dist", length(starts_ends))
write.table(file="data/process/start_end_samples.dist", distances[starts_ends, starts_ends], col.names=F, quote=FALSE, append=T)
system('mothur "#nmds(phylip=data/process/start_end_samples.dist)"')

axes_file <- "data/process/start_end_samples.nmds.axes"

axes <- read.table(file=axes_file, header=T, row.names=1)
start_end_axes <- axes[c(starts, ends),]
treatments <- sapply(rownames(start_end_axes), get_treatment)

symbols <- pch[treatments]

plot(NA, xlim=range(axes[,1]), ylim=range(axes[,2]), xlab="NMDS Axis 1",
        ylab="NMDS Axis 2")
arrows(x0=axes[starts,1], y0=axes[starts,2], x1=axes[ends,1], y1=axes[ends,2], col="gray", lwd=2, length=0.1)
points(start_end_axes, col=clrs[treatments], pch=symbols)
```

Let's see whether we can build a random forest to predict whether a sample comes
from the beginning or the end of the model.

```{r}
n_trees <- 20000

rabund <- get_rel_abund("data/process/ab_aomdss.trim.contigs.good.unique.good.filter.unique.precluster.pick.pick.pick.an.unique_list.0.03.subsample.shared", "DSS")

start_end_rabund <- rabund[starts_ends,]
filter_start_end_rabund <- filter_rabund(start_end_rabund, 0.0075)
period <- rep("end", nrow(filter_start_end_rabund))
period[sapply(rownames(filter_start_end_rabund), get_day) == 0] <- "begin"

rf <- randomForest(filter_start_end_rabund, factor(period), importance=TRUE, ntree=n_trees)

```

Looks pretty crappy. Here's the confusion matrix:

`r kable(rf$confusion)`

The out of bag error rate was `r round(100*rf$err.rate[n_trees,"OOB"], 1)`%.


How about if we do it based on the final samples and see whether there was
overlap with the OTUs we detected at the beginning of the model.

```{r}
n_trees <- 10000
threshold <- 0.015

rabund <- get_rel_abund("data/process/ab_aomdss.trim.contigs.good.unique.good.filter.unique.precluster.pick.pick.pick.an.unique_list.0.03.subsample.shared", "DSS")

end_rabund <- rabund[ends,]

end_tumor_counts <- tumor_counts[gsub("_d.*", "", rownames(end_rabund)),1]

end_rf <- get_forest(end_rabund, end_tumor_counts, threshold, n_trees=10000)

plot_forest_fit(end_tumor_counts, end_rf, end_rabund, sapply(rownames(end_rabund), get_treatment))
plot_importance(end_rf)
plot_top_features(end_tumor_counts,end_rf, end_rabund, sapply(rownames(end_rabund), get_treatment))
end_simple <- simplify_model(end_tumor_counts, end_rf, end_rabund)
```

This model does a pretty good job of predicting the number of tumors, but it
is pretty redundant with the mBio paper and the baseline model.


Let's try building a model to connect change in relative abundance between
baseline and the end point with the number of tumors


```{r}
in_both <- !is.na(starts) & !is.na(ends)
good_starts <- starts[in_both]
good_ends <- ends[in_both]

rabund_starts <- rabund[good_starts,]
rabund_ends <- rabund[good_ends,]

rabund_diffs <- rabund_ends - rabund_starts
decent_move <- apply(rabund_diffs, 2, mean) > 0.01
decent_rabund_diffs <- rabund_diffs[,decent_move]
```

This results in 4 OTUs - Ecoli, Bacteroidetes, Odoribacter, and
Beta-proteobacteria. Not sure what it really says...
