# Random Forest Modeling


**Purpose:** We want to model the number of tumors based on the microbiome using
regression via random forests


### Setup

```{r setup}
library(randomForest, quietly=TRUE)
library(knitr, quietly=TRUE)
sessionInfo()
```

```{r knitr_settings, echo=TRUE, eval=TRUE}
opts_chunk$set(dev = c("png", "pdf"))
opts_chunk$set(results = "hold")
opts_chunk$set(fig.show = "hold")
opts_chunk$set(warning=FALSE)
opts_chunk$set(fig.align="center")
```

Let's read in the shared file and convert the numbers to relative abundances...

```{r input_shared}
shared.file <- "data/process/ab_aomdss.trim.contigs.good.unique.good.filter.unique.precluster.pick.pick.pick.an.unique_list.0.03.subsample.shared"
shared <- read.table(file=shared.file, row.names=2, header=T, stringsAsFactors=FALSE)

#get rid of the first two columns, which are not needed
cols_to_remove <- c("label", "numOtus")
shared <- shared[, !colnames(shared)%in% cols_to_remove]

#remove those columns that are labeled as being samples that only received DSS
rows_to_remove <- grepl("DSSonly", rownames(shared))
shared <- shared[!rows_to_remove, ]

#calculate the number of sequences and then get out the relative abundance
n_seqs <- sum(shared[1,])
shared <- shared/n_seqs
```

Now we want to get the tumor count data

```{r tumor_counts}
tumor_count.file <- "data/process/tumor_counts.tsv"
tumor_counts <- read.table(file=tumor_count.file, header=T)
```



Let's get the metadata from the sample names

```{r metadata}
#The sample names go treatment_eartag_day, so let's split them on the '_'
#and then return the appropriate slot in the resulting vector
get_name_part <- function(sample_name, column){
    name_parts <- unlist(strsplit(sample_name, "_"))
    return(name_parts[column])
}

#treatment is in slot 1
get_treatment <- function(sample_name){
    get_name_part(sample_name, column=1)
}

#eartag is in slot 2
get_eartag <- function(sample_name){
    get_name_part(sample_name, column=2)
}

#eartag is in slot 3; neg indicates that the sample was prior to day 0;
#need to convert string to a number
get_day <- function(sample_name){
    text.day <- get_name_part(sample_name, column=3)
    text.day <- gsub("d", "", text.day)
    text.day <- gsub("neg", "-", text.day)
    text.day <- as.numeric(text.day)
}

treatment <- sapply(rownames(shared), get_treatment)
eartag <- sapply(rownames(shared), get_eartag)
day <- sapply(rownames(shared), get_day)
```


### Get the baseline samples

I want to build a model based on all of the day 0 data and then predict the
number of tumors based on the observed data. Let's identify those
samples that are from day 0:

```{r baseline_setup}
baseline <- grepl("_d0", rownames(shared))  #baseline samples are from day 0
treatment_baseline <- treatment[baseline]   #update the metadata
eartag_baseline <- eartag[baseline]
day_baseline <- day[baseline]

shared_baseline <- shared[baseline,]        #extract the baseline samples

mice_baseline <- paste(treatment_baseline, eartag_baseline, sep="_")
tumor_counts_baseline <- tumor_counts[mice_baseline,]   #get the tumor count data
```


### Find the best threshold

One question is how best to filter the OTU data. Let's see what happens when we
alter the minimum average relative abundance across all samples:

```{r filter_test, cache=TRUE}
threshold <- seq(0,0.03, 0.005)
n_trees <- 10000 #number of trees for random forest

rsq <- rep(0, length(threshold))    #vector to hold the Rsquared values
n_otus <- rep(0, length(threshold)) #number of OTUs for the threshold

for(i in 1:length(threshold)){
    #keep those OTUs that are more than threshold across all samples
    filter <- apply(shared_baseline, 2, mean) > threshold[i]
    shared_baseline_filtered <- shared_baseline[,filter]
    rf <- randomForest(tumor_counts_baseline ~ ., data=shared_baseline_filtered,
                                    importance=TRUE, proximity=TRUE, ntree=n_trees)
    rsq[i] <- rf$rsq[n_trees] #percent variance explained
    n_otus[i] <- ncol(shared_baseline_filtered)
}

opt_thresh <- threshold[which.max(rsq)]
opt_thresh_notus <- n_otus[which.max(rsq)]
opt_thresh_rsq <- max(rsq)


plot(rsq~threshold, pch=19, xlab="Average relative abundance", ylab="Rsquared", ylim=c(0.4, max(rsq)))
text(x=threshold, y=0.45, label=n_otus)
```

From this plot, it appears that the best threshold would be `r opt_thresh`
(Rsq=`r round(opt_thresh_rsq, 2)`) and this would give us `r opt_thresh_notus`
OTUs.



### Find the best number of trees

Ok, let's press on and make sure we're using a sufficient number of trees with
our threshold:


```{r ntrees_test, cache=TRUE}
threshold <- opt_thresh
n_trees <- seq(1000,30000,1000) #number of trees for random forest

tree_rsq <- rep(0, length(n_trees))  #vector to hold the Rsquared values

for(i in 1:length(n_trees)){
    #keep those OTUs that are more than threshold across all samples
    filter <- apply(shared_baseline, 2, mean) > threshold
    shared_baseline_filtered <- shared_baseline[,filter]
    rf <- randomForest(tumor_counts_baseline ~ ., data=shared_baseline_filtered,
                                    importance=TRUE, proximity=TRUE, ntree=n_trees[i])
    tree_rsq[i] <- rf$rsq[n_trees[i]] #percent variance explained
}

plot(tree_rsq~n_trees, pch=19, xlab="Number of trees", ylab="Rsquared")
```

We see that over a wide spread in the number of trees used to model the data
that there really isn't much variation if we use 1,000 or 30,000 trees (min:
`r round(min(tree_rsq), 2)`, max: `r round(max(tree_rsq), 2)`). We'll stick with
10,000 trees for the rest of this analysis.




### Run the model

We'll run the model with 10,000 trees and use `r opt_thresh` as our
threshold for filtering the data:

```{r get_model, cache=TRUE}
n_trees <- 10000
threshold <- opt_thresh

#filter the data
filter <- apply(shared_baseline, 2, mean) > threshold
shared_baseline_filtered <- shared_baseline[,filter]

#build the model
rf_baseline <- randomForest(tumor_counts_baseline ~ ., data=shared_baseline_filtered,
                                        importance=TRUE, proximity=TRUE, ntree=n_trees)

#get the Rsquared value
rf_baseline_rsq <- rf_baseline$rsq[n_trees] #percent variance explained

#get the importance data
rf_baseline_importance <- importance(rf_baseline)   #extract the importance data

#plot the error over the forest
plot(rf_baseline, main="")
```

Looks like the model has stabilized at `r n_trees`. Let's see the importance
plots:

```{r importance_plot}
#plot the importance data
varImpPlot(rf_baseline, main="")
```

This model results in an Rsq value of `r round(rf_baseline_rsq, 2)`. Inspection
of the Gini coefficient plot suggests that there are three OTUs that stick out
as being highly informative, followed by an additional three OTUs that are also
informative.



### Simplify the model

Let's sort the OTUs by their Gini coefficient and build a model for all of the
OTUs.


```{r simplify_model, cache=TRUE}
sorted_importance <- rf_baseline_importance[order(rf_baseline_importance[,2], decreasing=T),]

n_otus <- nrow(sorted_importance)
rf_simplify_rsq <- rep(0, n_otus)

for(i in 2:n_otus){
    simple_shared <- shared_baseline_filtered[,colnames(shared_baseline_filtered)
                                                    %in% rownames(sorted_importance)[1:i]]
    rf_simplify <- randomForest(tumor_counts_baseline ~ ., data=simple_shared,
                                            importance=TRUE, ntree=n_trees)
    rf_simplify_rsq[i] <- rf_simplify$rsq[n_trees] #percent variance explained
}

plot(rf_simplify_rsq[-1]~c(2:n_otus), ylab="Rsquared", xlab="Number of OTUs in model")
```

Here we see that the model does pretty well with just a few parameters.



### Fit data to model

Let's compare the number of predicted tumors to the number of observed tumors
based on the full model.

```{r model_fit}
rf_baseline_pred <- predict(rf_baseline, shared_baseline_filtered)

max_value <- max(c(rf_baseline_pred,tumor_counts_baseline))

clrs <- c(Metro = "red",
    VancStrep = "red",
    Strep = "darkgreen",
    VancMetro = "darkgreen",
    Vanc = "blue",
    StrepMetro = "blue",
    AllAbs = "black",
    NoAbs = "black")

pch <- c(Metro = 19,
    VancStrep = 1,
    Strep = 19,
    VancMetro = 1,
    Vanc = 19,
    StrepMetro = 1,
    AllAbs = 19,
    NoAbs = 1)

labels <- c(NoAbs = "No antibiotics",
    AllAbs = "All antibiotics",
    Metro = "Metronidozole",
    VancStrep = "delMetronidozole",
    Strep = "Streptomycin",
    VancMetro = "delStreptomycin",
    Vanc = "Vancomycin",
    StrepMetro = "delVancomycin")


plot(rf_baseline_pred~tumor_counts_baseline, pch=pch[treatment_baseline],
    col=clrs[treatment_baseline], ylim=c(0,max_value), xlim=c(0,max_value),
    cex=1.5, xlab="Observed number of tumors", ylab="Predicted number of tumors")
legend(x=0, y=max_value, legend=labels, pch=pch[names(labels)],
        col=clrs[names(labels)])
```


### Plot the top six features with tumor counts

Let's generate a plot with the number of tumors on the y-axis and the OTU's
relative abundance on the x-axis:

```{r rabund_tumor_plot, fig.width=7.5, fig.height=5}

tax <- read.table(file="data/process/ab_aomdss.trim.contigs.good.unique.good.filter.unique.precluster.pick.pick.pick.an.unique_list.0.03.cons.taxonomy", header=T, row.names=1)

tax$Taxonomy <- gsub("unclassified.*", "", tax$Taxonomy)
tax$Taxonomy <- gsub("\\(\\d*\\);$", "", tax$Taxonomy)
tax$Taxonomy <- gsub(".*;", "", tax$Taxonomy)

otus <- rownames(sorted_importance)[1:6]
pretty_otus <- gsub("Otu0*", "OTU ", otus)
otu_labels <- paste0("(", pretty_otus, ")", "\nGini: ", round(sorted_importance[1:6,2], 1))
otu_labels <- paste(tax[otus,2], otu_labels, sep=" ")

shared_nozeros <- shared_baseline_filtered
#shared_nozeros[shared_baseline_filtered==0] <- 5/25000

par_orig <- par()$mar
par(mar=c(5,4,0.5,0.5))
layout(matrix(1:6, nrow=2, byrow=T))
for(i in 1:6){

    row <- ifelse(i<=3,1,2)
    column <- ifelse(i<=3, i, i-3)

    xlab <- ""
    if(row==2){
        xlab <- "Relative abundance (%)"
    }

    ylab <- ""
    if(column==1){
        ylab <- "Number of tumors"
    }
    rel_abund <- shared_nozeros[,otus[i]]

    plot(rel_abund,tumor_counts_baseline, log="x", pch=pch[treatment_baseline],
        col=clrs[treatment_baseline], ylab=ylab, xlab=xlab,
        xlim=c(1e-4, 1), ylim=c(0,27), yaxt="n",
        xaxt="n", cex.lab=1.5)

    zeroes <- rel_abund == 0
    x_zeroes <- runif(sum(zeroes),1.0e-4,1.5e-4)#rep(1.2e-4, sum(zeroes))
    points(x=x_zeroes, tumor_counts_baseline[zeroes], pch=pch[treatment_baseline[zeroes]],
            col=clrs[treatment_baseline[zeroes]])
    abline(v=2.2e-4, col="gray")

    text(x=0.8e-4, y=25, label=otu_labels[i], pos=4, font=2)

    if(row == 2){
        axis(1, at=c(1.25e-4, 1e-3,1e-2,1e-1,1),
                label=c("0", "0.1", "1", "10", "100"),
                cex.axis=1.5)
    }
    if(column == 1){
        axis(2, las=2, cex.axis=1.5)
    }
}

par(mar=par_orig)
```

These plots highlight a few important results:
* Anaeroplasma, Porphyromonadaceae, and Prevotella increase with tumor count
* Enterobacteriaceae, Ureaplasma, and Lactobacillus decrease with tumor count
