# Random Forest Modeling


**Purpose:** We want to model the number of tumors based on the microbiome using
regression via random forests


### Setup

```{r setup, results = "hold"}
library(randomForest, quietly=TRUE)
library(knitr, quietly=TRUE)
sessionInfo()

set.seed(6201976)
source("code/rf_baseline_analysis.R")
```

```{r knitr_settings, echo=TRUE, eval=TRUE}
opts_chunk$set(dev = c("png", "pdf"))
opts_chunk$set(results = "hold")
opts_chunk$set(fig.show = "hold")
opts_chunk$set(warning = FALSE)
opts_chunk$set(fig.align = "center")
opts_chunk$set(echo = FALSE)
opts_chunk$set(cache = FALSE)
```

Let's get the relative abundance of the OTUs in our samples at baseline (Day 0).
To do this, we will read in the shared file and convert the numbers to relative
abundances and get the tumor count data. Using the sample names we will also
extract the treatment, mouse eartag number, and the day of the experiment.


```{r input_shared}
shared_file <- "data/process/ab_aomdss.trim.contigs.good.unique.good.filter.unique.precluster.pick.pick.pick.an.unique_list.0.03.subsample.shared"
rel_abund <- get_rel_abund(paste0(PROJHOME, "/", shared_file), "DSS")

treatment <- sapply(rownames(rel_abund), get_treatment)
eartag <- sapply(rownames(rel_abund), get_eartag)
day <- sapply(rownames(rel_abund), get_day)

baseline <- day == 0                        #baseline samples are from day 0
treatment_baseline <- treatment[baseline]   #update the metadata
eartag_baseline <- eartag[baseline]         #update the metadata
day_baseline <- day[baseline]               #update the metadata
rabund_baseline <- rel_abund[baseline,]     #extract the baseline samples

#what mice do we have baseline data for? necessary because the tumor data
#is indexed by `treatment_eartag`
mice_baseline <- paste(treatment_baseline, eartag_baseline, sep="_")

tumor_count.file <- paste0(PROJHOME, "/data/process/tumor_counts.tsv")
tumor_counts <- read.table(file=tumor_count.file, header=T)
tumor_counts_baseline <- tumor_counts[mice_baseline,]   #get the tumor count data
```

### Find the best forest

One question is how best to filter the OTU data. Let's see what happens when we
alter the minimum average relative abundance across all samples:

```{r filter_test}
threshold <- seq(0,0.03, 0.005)
n_trees <- 10000 #number of trees for random forest

rf_thresholds <- lapply(X=threshold, FUN=get_forest, rabund=rabund_baseline,
                         dependent=tumor_counts_baseline, n_trees=n_trees)

thresh_rsq <- sapply(rf_thresholds, get_rsq)

max_index <- which.max(thresh_rsq)
rf_baseline_forest <- rf_thresholds[[max_index]]    #get the forest
rf_baseline_thresh <- threshold[max_index]          #get the threshold that was used

#get filtered rabund table
rf_baseline_rabund <- filter_rabund(rabund_baseline, rf_baseline_thresh)

plot(thresh_rsq~threshold, pch=19, xlab="Average relative abundance",
                                ylab="Rsquared", ylim=c(0.4, max(thresh_rsq)))
text(x=threshold, y=0.45, label=sapply(rf_thresholds, get_n_otus))
```

From this plot, it appears that the best threshold would be
`r rf_baseline_thresh` (Rsq=`r round(get_rsq(rf_baseline_forest), 2)`) and this would
give us `r get_n_otus(rf_baseline_forest)` OTUs. Let's press on and make sure we're
using a sufficient number of trees with our threshold:

```{r ntrees_test}
plot(rf_baseline_forest, main="")
```

Looks like the model has stabilized at `r n_trees` trees. Let's see the
importance plots:

```{r importance_plot, tidy=TRUE}
par(mar=c(5,5,0,0))
varImpPlot(rf_baseline_forest, main="")
```

Inspection of the Gini coefficient plot (right side of above plot) suggests that
there are three OTUs that stick out as being highly informative, followed by an
additional three OTUs that are also informative.


### Simplify the model

Let's sort the OTUs by their Gini coefficient and build models where we add each
OTU in a stepwise manner.


```{r simplify_model}
rf_simplify_rsq <- simplify_model(tumor_counts_baseline, rf_baseline_forest, rabund_baseline)
plot(rf_simplify_rsq[-1]~c(2:length(rf_simplify_rsq)), ylab="Rsquared", xlab="Number of OTUs in model")
```

Thus, we see that the model does pretty well with just a few parameters.


### Fit data to model

Let's compare the number of predicted tumors to the number of observed tumors
based on the full model.

```{r model_fit}
plot_forest_fit(tumor_counts_baseline, rf_baseline_forest, rf_baseline_rabund, treatment_baseline)
```

Pretty sweet!


### Plot the top six features with tumor counts

Let's generate a plot with the number of tumors on the y-axis and the OTU's
relative abundance on the x-axis:

```{r rabund_tumor_plot, fig.width=7.5, fig.height=5}
plot_top_features(tumor_counts_baseline, sorted_importance, rf_baseline_rabund, treatment_baseline)
```

These plots highlight a few important results:
* Anaeroplasma, Porphyromonadaceae, and Prevotella increase with tumor count
* Enterobacteriaceae, Ureaplasma, and Lactobacillus decrease with tumor count
